---
title: "data_sci_II_midterm_v1"
author: "Zihan Xiong"
date: "2026-03-27"
output: html_document
---

---
title: "Data Science II Midterm Project"
author: "Midterm Project Group 29"
date: "2026-03-24"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: cosmo
---
```{r,warning=FALSE, message=FALSE,echo=FALSE}
library(tidyverse)
library(caret)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(corrplot)
library(glmnet)

library(mgcv)
library(earth)

set.seed(123)
```

# **Preparation and Inspection**  
## Loading Data  
```{r,warning=FALSE, message=FALSE,echo=FALSE}
load("data/datA.RData")
load("data/datB.RData")

str(datA)
str(datB)
```

## Set variable type, dimension, and data summary
```{r,warning=FALSE, message=FALSE,echo=FALSE}
datA <- datA |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

datB <- datB |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

dim(datA); summary(datA)
dim(datB); summary(datB)
```

## Check for Missing Data   
```{r,warning=FALSE, message=FALSE,echo=FALSE}
colSums(is.na(datA))
colSums(is.na(datB)) # Check if any data missing.
```  

# **Part 1: PFS model (1990 - 1995)**  

## *○ Data Partition*

```{r,warning=FALSE, message=FALSE,echo=FALSE}
set.seed(123)
# partition first, 0.8
datsplit <- initial_split(datA, prop = 0.8)
train <- training(datsplit) |> 
  select(-id)
test <- testing(datsplit) |> 
  select(-id)
```

### Data Visualization and featurePlot

```{r,warning=FALSE, message=FALSE,echo=FALSE}
# visualization using train data
# continuous: age, bmi, tumor_size, lymph_nodes, Ki67
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x = train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
            y = train$log_PFS,
            plot = "scatter",
            span = 0.5,
            labels = c("Predictors", "log_PFS"),
            type = c("p", "smooth"),
            layout=c(3,2))
```

While looking at the features plots, we can observe that:

`lymph_nodes`: negative relationship with log_PFS — more positive lymph nodes, shorter PFS. The trend appears approximately linear.

`Ki67`: negative relationship with log_PFS. Higher proliferation index associated with shorter PFS, though the trend is relatively mild.

`age`: very weak relationship. Smooth curve is flat.

`bmi`: slight negative relationship. 

`tumor_size`: ngeative relationship. larger tumor size associated with shorter PFS.

```{r,warning=FALSE, message=FALSE,echo=FALSE}
# also take a look at the Pearson correlation coefficient
cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")], 
    train$log_PFS)
```

```{r,warning=FALSE, message=FALSE,echo=FALSE}
# corrplot, none r<0.8
cor_matrix <- cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")])
corrplot(cor_matrix, method = "number", type = "upper")
```

```{r,warning=FALSE, message=FALSE,echo=FALSE}
# categorical: stage, tumor_grade, ER, PR, HER2, menopause, surgery, chemo, hormonal_therapy
# I'm using box plot instead of the feature plot
p1 <- ggplot(train, aes(x = stage, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p2 <- ggplot(train, aes(x = tumor_grade, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p3 <- ggplot(train, aes(x = ER, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p4 <- ggplot(train, aes(x = PR, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p5 <- ggplot(train, aes(x = HER2, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p6 <- ggplot(train, aes(x = menopause, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p7 <- ggplot(train, aes(x = surgery, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p8 <- ggplot(train, aes(x = chemo, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p9 <- ggplot(train, aes(x = hormonal_therapy, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

Notice that:

`stage`: Stage 1 patients having the highest log_PFS and Stage 3 the lowest (F = 81.0, p < 0.001), more advanced disease stage is associated with shorter PFS.

`tumor_grade`: Grade 1 patients had notably higher log_PFS, poorer tumor differentiation is associated with worse prognosis.

`ER`: ER-positive patients had substantially higher log_PFS than ER-negative patients (F = 138.4, p < 0.001)

`hormonal_therapy`: Patients receiving hormonal therapy had markedly higher log_PFS (F = 89.6, p < 0.001)

### ANOVA  
```{r,warning=FALSE, message=FALSE,echo=FALSE}
cat_vars <- c("stage", "tumor_grade", "ER", "PR", "HER2", 
              "menopause", "surgery", "chemo", "hormonal_therapy")

anova_results <- data.frame(
  Variable = cat_vars,
  F_value = NA,
  P_value = NA
)

for(i in seq_along(cat_vars)){
  fit <- aov(log_PFS ~ train[[cat_vars[i]]], data = train)
  f_val <- summary(fit)[[1]]$`F value`[1]
  p_val <- summary(fit)[[1]]$`Pr(>F)`[1]
  anova_results$F_value[i] <- round(f_val, 3)
  anova_results$P_value[i] <- round(p_val, 4)
}

anova_results
```

## resampling and re-build the models
```{r,warning=FALSE, message=FALSE,echo=FALSE}
set.seed(123)
folds <- createMultiFolds(train$log_PFS, k = 10, times = 5)

ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = folds
)

# LM

lm_fit <- train(log_PFS ~ ., data = train, method = "lm", metric = "RMSE", trControl = ctrl)

# Ridge

ridge_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(-5, 5, length = 100))),
                   trControl = ctrl)

# Lasso

lasso_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(-5, 2, length = 100))),
                   trControl = ctrl)

# Elastic Net

enet_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                         lambda = exp(seq(-5, 5, length = 100))),
                  trControl = ctrl)

# GAM 

gam_fit <- train(log_PFS ~ ., data = train, method = "gam",  metric = "RMSE", trControl = ctrl)

# MARS

mars_fit <- train(log_PFS ~ ., data = train, method = "earth", metric = "RMSE",
                  tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                  trControl = ctrl)
# compare
resamp <- resamples(list(LM = lm_fit, Ridge = ridge_fit, Lasso = lasso_fit,
                         ENet = enet_fit, GAM = gam_fit, MARS = mars_fit))
summary(resamp)
bwplot(resamp, metric = "RMSE")
```




## Model selection based on the resampling results

If we choose to use GAM as the final model:
```{r,warning=FALSE, message=FALSE,echo=FALSE}
# GAM as final model
summary(gam_fit$finalModel)
plot(gam_fit$finalModel, pages = 1)

# Test set evaluation
gam_pred <- predict(gam_fit, newdata = test)
gam_test_rmse <- sqrt(mean((test$log_PFS - gam_pred)^2))
gam_test_r2 <- cor(test$log_PFS, gam_pred)^2
gam_test_rmse
gam_test_r2
```

## Generalization GAM on datB
```{r}
## generalization on datB
# GAM trained on datA, predict on entire datB
gam_pred_datB <- predict(gam_fit, newdata = datB)

gam_rmse_datB <- sqrt(mean((datB$log_PFS - gam_pred_datB)^2))
gam_r2_datB <- cor(datB$log_PFS, gam_pred_datB)^2

gam_rmse_datB
gam_r2_datB
```

## split
```{r}
set.seed(123)
datsplit_b <- initial_split(datB, prop = 0.8)
train_b <- training(datsplit_b) |> select(-id)
test_b  <- testing(datsplit_b) |> select(-id)
```

## re-build models on datB
```{r}
# resampling setup
set.seed(123)
folds_b <- createMultiFolds(train_b$log_PFS, k = 10, times = 5)
ctrl_b <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = folds_b
)

# LM
lm_fit_b <- train(log_PFS ~ ., data = train_b, method = "lm", metric = "RMSE", trControl = ctrl_b)
# Ridge
ridge_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                     tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(-5, 5, length = 50))),
                     trControl = ctrl_b)
# Lasso
lasso_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                     tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(-5, 2, length = 50))),
                     trControl = ctrl_b)
# Elastic Net
enet_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                    tuneGrid = expand.grid(alpha = seq(0, 1, length = 11),
                                           lambda = exp(seq(-5, 5, length = 50))),
                    trControl = ctrl_b)

# GAM 
gam_fit_b <- train(log_PFS ~ ., data = train_b, method = "gam", metric = "RMSE", trControl = ctrl_b)
# MARS
mars_fit_b <- train(log_PFS ~ ., data = train_b, method = "earth", metric = "RMSE",
                    tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                    trControl = ctrl_b)

# compare
resamp_b <- resamples(list(LM = lm_fit_b, Ridge = ridge_fit_b, Lasso = lasso_fit_b,
                           ENet = enet_fit_b, GAM = gam_fit_b, MARS = mars_fit_b))
summary(resamp_b)
bwplot(resamp_b, metric = "RMSE")
```

## GAM comparison across cohorts
```{r}
# datA GAM summary and smooth curves
summary(gam_fit$finalModel)
plot(gam_fit$finalModel, pages = 1)

# datB GAM summary and smooth curves
summary(gam_fit_b$finalModel)
plot(gam_fit_b$finalModel, pages = 1)

# datB test set evaluation
gam_pred_b <- predict(gam_fit_b, newdata = test_b)
gam_test_rmse_b <- sqrt(mean((test_b$log_PFS - gam_pred_b)^2))
gam_test_r2_b <- cor(test_b$log_PFS, gam_pred_b)^2
gam_test_rmse_b
gam_test_r2_b
```

GAM before generalization RMSE: 0.840
after: 0.731265


## Model selection based on the resampling results

If we want to choose LASSO as our final model:

```{r}
# if we choose to use LASSO as the final model (because of easier interpretation)
lasso_pred <- predict(lasso_fit, newdata = test)
lasso_test_rmse <- sqrt(mean((test$log_PFS - lasso_pred)^2))
lasso_test_r2 <- cor(test$log_PFS, lasso_pred)^2

lasso_test_rmse
lasso_test_r2

# Lasso coefficients
coef(lasso_fit$finalModel, lasso_fit$bestTune$lambda)

```

## generalization on datB
```{r}
# if we select LASSO
lasso_pred_datB <- predict(lasso_fit, newdata = datB)

lasso_rmse_datB <- sqrt(mean((datB$log_PFS - lasso_pred_datB)^2))
lasso_r2_datB <- cor(datB$log_PFS, lasso_pred_datB)^2

lasso_rmse_datB
lasso_r2_datB
```

## split
```{r}
set.seed(123)
datsplit_b <- initial_split(datB, prop = 0.8)
train_b <- training(datsplit_b) |> select(-id)
test_b  <- testing(datsplit_b) |> select(-id)
```

## re-build models on datB
```{r}
# resampling setup
set.seed(123)
folds_b <- createMultiFolds(train_b$log_PFS, k = 10, times = 5)
ctrl_b <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = folds_b
)

# LM
lm_fit_b <- train(log_PFS ~ ., data = train_b, method = "lm", metric = "RMSE", trControl = ctrl_b)
# Ridge
ridge_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                     tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(-5, 5, length = 50))),
                     trControl = ctrl_b)
# Lasso
lasso_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                     tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(-5, 2, length = 50))),
                     trControl = ctrl_b)
# Elastic Net
enet_fit_b <- train(log_PFS ~ ., data = train_b, method = "glmnet", metric = "RMSE",
                    tuneGrid = expand.grid(alpha = seq(0, 1, length = 11),
                                           lambda = exp(seq(-5, 5, length = 50))),
                    trControl = ctrl_b)

# GAM 
gam_fit_b <- train(log_PFS ~ ., data = train_b, method = "gam", metric = "RMSE", trControl = ctrl_b)
# MARS
mars_fit_b <- train(log_PFS ~ ., data = train_b, method = "earth", metric = "RMSE",
                    tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                    trControl = ctrl_b)

# compare
resamp_b <- resamples(list(LM = lm_fit_b, Ridge = ridge_fit_b, Lasso = lasso_fit_b,
                           ENet = enet_fit_b, GAM = gam_fit_b, MARS = mars_fit_b))
summary(resamp_b)
bwplot(resamp_b, metric = "RMSE")
```

## coefficient of LASSO
```{r}
# datA coefficient lasso
coef(lasso_fit$finalModel, lasso_fit$bestTune$lambda)
# datB coefficient lasso
coef(lasso_fit_b$finalModel, lasso_fit_b$bestTune$lambda)
```

## coefficient of GAM
```{r}
# automatically non-linear variables by caret
# datA
summary(gam_fit$finalModel)
plot(gam_fit$finalModel)
#datB
summary(gam_fit_b$finalModel)
plot(gam_fit_b$finalModel)
```

懒得打英语了hhh,可以对比一下lasso和gam,共同点:
1. HER2效应消失
2. bmi的非线性
我觉得重点我们可以解释模型的前后coefficient的变化

