---
title: "Data Science II Midterm Project"
author: "Midterm Project Group 29"
date: "2026-03-24"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: cosmo
---
```{r,warning=FALSE, message=FALSE}
library(tidyverse)
library(caret)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(corrplot)
library(glmnet)

library(mgcv)
library(earth)
library(kableExtra)

dir.create("figures", showWarnings = FALSE)
```

# **Preparation and Inspection**  
## Loading Data  
```{r}
load("data/datA.RData")
load("data/datB.RData")

str(datA)
str(datB)
```

## Set variable type, dimension, and data summary
```{r}
datA <- datA |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  ) |> 
  select(-id)

datB <- datB |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  ) |> 
  select(-id)

dim(datA); summary(datA)
dim(datB); summary(datB)
```
## Check for Missing Data   
```{r}
colSums(is.na(datA))
colSums(is.na(datB))
```   


# **Part 1: PFS model (1990 - 1995)**  

## *○ Data Partition*

```{r}
set.seed(123)
# partition first, 0.8
datsplit <- initial_split(datA, prop = 0.8)
train <- training(datsplit)
test <- testing(datsplit)
```

### Data Visualization and featurePlot

```{r,warning=FALSE}
# visualization using train data
# continuous: age, bmi, tumor_size, lymph_nodes, Ki67
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

fp1 <- featurePlot(x = train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
            y = train$log_PFS,
            plot = "scatter",
            span = 0.5,
            labels = c("Predictors", "log_PFS"),
            type = c("p", "smooth"),
            layout=c(3,2))
fp1
```

While looking at the features plots, we can observe that:

`lymph_nodes`: negative relationship with log_PFS — more positive lymph nodes, shorter PFS. The trend appears approximately linear.

`Ki67`: negative relationship with log_PFS. Higher proliferation index associated with shorter PFS, though the trend is relatively mild.

`age`: very weak relationship. Smooth curve is flat.

`bmi`: slight nonlinear, weak relationship. 

`tumor_size`: ngeative relationship. larger tumor size associated with shorter PFS.

As `bmi` shows the possibility of nonlinear relationships, flexible models such as GAM and MARS were considered as candidate models.

```{r}
# Pearson correlation with log_PFS
cor_with_pfs <- cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
    train$log_PFS)
cor_df <- data.frame(Variable = rownames(cor_with_pfs),
                     Correlation = round(cor_with_pfs[, 1], 4))
kbl_cor <- kbl(cor_df, caption = "Pearson Correlation with log_PFS",
               row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_cor
```
Pearson correlation suggested moderate negative linear associations of tumor size, lymph nodes, and Ki67 with `log_PFS`, while age and BMI showed little linear correlation. As Pearson correlation captures only linear relationships, the near-zero correlation for BMI does not rule out a nonlinear association.

```{r}
# corrplot, none r<0.8
cor_matrix <- cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")])
corrplot(cor_matrix, method = "number", type = "upper")
```

Tumor size and lymph node count showed moderate-to-high correlation (r ≈ 0.70), suggesting potential multicollinearity. However, both variables are clinically meaningful and were retained in the model.

```{r}
# categorical: stage, tumor_grade, ER, PR, HER2, menopause, surgery, chemo, hormonal_therapy
# I'm using box plot instead of the feature plot
p1 <- ggplot(train, aes(x = stage, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p2 <- ggplot(train, aes(x = tumor_grade, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p3 <- ggplot(train, aes(x = ER, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p4 <- ggplot(train, aes(x = PR, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p5 <- ggplot(train, aes(x = HER2, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p6 <- ggplot(train, aes(x = menopause, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p7 <- ggplot(train, aes(x = surgery, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p8 <- ggplot(train, aes(x = chemo, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p9 <- ggplot(train, aes(x = hormonal_therapy, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
bp_combined <- (p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
bp_combined
```

Notice that:

`stage`: Stage 1 patients having the highest log_PFS and Stage 3 the lowest (F = 81.0, p < 0.001), more advanced disease stage is associated with shorter PFS.

`tumor_grade`: Grade 1 patients had notably higher log_PFS, poorer tumor differentiation is associated with worse prognosis.

`ER`: ER-positive patients had substantially higher log_PFS than ER-negative patients (F = 138.4, p < 0.001)

`hormonal_therapy`: Patients receiving hormonal therapy had markedly higher log_PFS (F = 89.6, p < 0.001)

### ANOVA  
```{r}
cat_vars <- c("stage", "tumor_grade", "ER", "PR", "HER2", 
              "menopause", "surgery", "chemo", "hormonal_therapy")

anova_results <- data.frame(
  Variable = cat_vars,
  F_value = NA,
  P_value = NA
)

for(i in seq_along(cat_vars)){
  fit <- aov(log_PFS ~ train[[cat_vars[i]]], data = train)
  f_val <- summary(fit)[[1]]$`F value`[1]
  p_val <- summary(fit)[[1]]$`Pr(>F)`[1]
  anova_results$F_value[i] <- round(f_val, 3)
  anova_results$P_value[i] <- round(p_val, 4)
}

kbl_anova <- kbl(anova_results, digits = 4,
                 caption = "One-Way ANOVA: Categorical Predictors vs log_PFS") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_anova

```   
Menopausal status was retained in the model due to its clinical relevance, even though it was not statistically significant in the univariate analysis.

## *○ Model Selection*   
Since the response variable, log-transformed PFS, is continuous, a linear regression model was considered appropriate. Exploratory data analysis showed that most continuous predictors had approximately linear relationships with log_PFS, with only mild non-linearity observed for BMI. Therefore, linear regression was used as the primary modeling approach.

In addition, given the evidence of nonlinear effects between log(PFS) and BMI, flexible models such as GAM and MARS were also considered as candidate models.

### Base model (linear)   
```{r, linear regression 1}
set.seed(123)

lm_fit <- lm(log_PFS ~ . , data = train)
summary(lm_fit)

kbl_lm <- kbl(tidy(lm_fit), digits = 4,
              caption = "Full Linear Regression Model Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```
The multiple linear regression model was statistically significant (p < 0.001), with an R² of `r summary(lm_fit)$r.squared` and adjusted R² of `r summary(lm_fit)$adj.r.squared`, indicating that approximately `r round(summary(lm_fit)$r.squared,2)*100`% of the variability in log-transformed PFS was explained by the predictors.

### Resampling and Alternative models   
Considering the potential limitations of linear models in predictive performance, and given that exploratory analysis suggested the presence of nonlinear relationships among some variables, more flexible models such as GAM and MARS were introduced to capture potential nonlinear effects. 

In addition, given the possible correlations among predictors, regularization methods including Ridge, LASSO, and Elastic Net were further considered to improve model stability and generalization. 

Finally, all candidate models were evaluated and compared using cross-validated RMSE implemented via the caret package.

#### Cross-validation setup
```{r, Cross validation setup}
set.seed(123)
folds <- createMultiFolds(train$log_PFS, k = 10, times = 5)

ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = folds
)
```

#### Linear Model and Regularized Models
```{r, Linear}
set.seed(123)
# LM

lm_fit <- train(log_PFS ~ ., data = train, method = "lm", metric = "RMSE", trControl = ctrl)

# Ridge

ridge_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(-5, 5, length = 100))),
                   trControl = ctrl)

# Lasso

lasso_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(-5, 2, length = 100))),
                   trControl = ctrl)

# Elastic Net

enet_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                         lambda = exp(seq(-5, 5, length = 100))),
                  trControl = ctrl)
lasso_coefs <- coef(lasso_fit$finalModel, lasso_fit$bestTune$lambda)
lasso_df <- data.frame(
  Term = rownames(lasso_coefs),
  Coefficient = round(as.numeric(lasso_coefs), 6)
) |> filter(Coefficient != 0)

kbl_lasso <- kbl(lasso_df, caption = "LASSO Non-Zero Coefficients",
                 row.names = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_lasso

summary(lm_fit$finalModel)
```

#### Nonlinear Models
```{r, Nonlinear}
set.seed(123)
# GAM 

gam_fit <- train(log_PFS ~ ., data = train, method = "gam",  metric = "RMSE", trControl = ctrl)

# MARS

mars_fit <- train(log_PFS ~ ., data = train, method = "earth", metric = "RMSE",
                  tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                  trControl = ctrl)
```

### RMSE results
```{r}
resamp <- resamples(list(LM = lm_fit, Ridge = ridge_fit, Lasso = lasso_fit,
                         ENet = enet_fit, GAM = gam_fit, MARS = mars_fit))
resamp_summary <- summary(resamp)

# Save RMSE summary table
rmse_stats <- as.data.frame(resamp_summary$statistics$RMSE)
rmse_stats$Model <- rownames(rmse_stats)
rmse_stats <- rmse_stats |> select(Model, everything())
kbl_rmse <- kbl(rmse_stats, digits = 4,
                caption = "Cross-Validated RMSE Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_rmse


bw <- bwplot(resamp, metric = "RMSE")
bw
```

The predictive performance of all candidate models was highly comparable, with only minimal differences in RMSE. Although GAM achieved the lowest RMSE, the improvement over the linear model was negligible (<1%), suggesting that increased model flexibility did not lead to meaningful gains in predictive performance, the linear regression model already captures most of the signal. 

## *○ Final model*   
Given the similar performance and the advantage of interpretability, the linear model was selected as the final model.
```{r}
set.seed(123)
lm_fit_datA <- lm(log_PFS ~ . , data = train)

lm_pred <- predict(lm_fit_datA, newdata = test)

lmA_rmse <- sqrt(mean((test$log_PFS - lm_pred)^2))
lmA_r2 <- cor(test$log_PFS, lm_pred)^2

# Final model summary table
kbl_final <- kbl(tidy(lm_fit_datA), digits = 4,
                 caption = "Final Linear Model (Part I) Coefficients") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_final


# Test performance table
perf_A <- data.frame(Metric = c("RMSE", "R-squared"),
                     Value = round(c(lmA_rmse, lmA_r2), 4))
kbl_perf <- kbl(perf_A, caption = "Final Model Test Performance (Part I)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_perf

```

```{r}
# residuals plot
plot(lm_pred, test$log_PFS - lm_pred,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted (Part I)")
abline(h = 0, col = "red")

# QQ plot
qqnorm(test$log_PFS - lm_pred, main = "Q-Q Plot of Residuals (Part I)")
qqline(test$log_PFS - lm_pred, col = "red")
```
Residuals are randomly scattered around zero with no clear pattern, suggesting no major violation of linearity or constant variance.   
The Q-Q plot suggests that residuals are approximately normally distributed, with only minor deviations in the tails.   
These diagnostic results suggest that the model assumptions are reasonably satisfied, indicating that the fitted linear model is adequate.  

```{r}
test_stats <- data.frame(
  Statistic = c("Mean", "SD", "Min", "Max"),
  Value = round(c(mean(test$log_PFS), sd(test$log_PFS),
                  min(test$log_PFS), max(test$log_PFS)), 4)
)
kbl_stats <- kbl(test_stats,
                 caption = "Test Set log_PFS Summary Statistics") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_stats

```


# **Part 2: PFS model (2000 - 2005)**  
## *○ Generalization lm on datB*

```{r, warning=FALSE}
set.seed(123)
pred_b <- predict(lm_fit, newdata = datB)
lm_rmse_datB <- sqrt(mean((datB$log_PFS - pred_b)^2))
lm_r2_datB <- cor(datB$log_PFS, pred_b)^2

gen_perf <- data.frame(
  Dataset = c("datA (test)", "datB (generalization)"),
  RMSE = round(c(lmA_rmse, lm_rmse_datB), 4),
  R_squared = round(c(lmA_r2, lm_r2_datB), 4)
)
kbl_gen <- kbl(gen_perf, caption = "Generalization Performance: datA Model on datB") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_gen

```

The model shows reduced performance when applied to the new dataset, as indicated by an increase in RMSE (from `r lmA_rmse` to `r lm_rmse_datB`) and a decrease in R-squared (from `r lmA_r2` to `r lm_r2_datB`).

This suggests that the relationships between predictors and progression-free survival may differ between the two cohorts.

## *○ Refiting Model*  

```{r}
set.seed(123)
datsplit_b <- initial_split(datB, prop = 0.8)
train_b <- training(datsplit_b)
test_b  <- testing(datsplit_b)
```

```{r,warning=FALSE}
# visualization using train data
# continuous: age, bmi, tumor_size, lymph_nodes, Ki67
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

fp2 <- featurePlot(x = train_b[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
            y = train_b$log_PFS,
            plot = "scatter",
            span = 0.5,
            labels = c("Predictors", "log_PFS"),
            type = c("p", "smooth"),
            layout = c(3, 2))
fp2
```

```{r}
set.seed(123)

lm_fit_datB <- lm(log_PFS ~ . , data = train_b)
summary(lm_fit_datB)

pred_b <- predict(lm_fit_datB, newdata = test_b)
lmB_rmse <- sqrt(mean((test_b$log_PFS - pred_b)^2))
lmB_r2 <- cor(test_b$log_PFS, pred_b)^2

# Part 2 model summary table
kbl_lmB <- kbl(tidy(lm_fit_datB), digits = 4,
               caption = "Refitted Linear Model (Part II) Coefficients") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_lmB


# Part 2 test performance
perf_B <- data.frame(Metric = c("RMSE", "R-squared"),
                     Value = round(c(lmB_rmse, lmB_r2), 4))
kbl_perfB <- kbl(perf_B, caption = "Refitted Model Test Performance (Part II)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_perfB

```
Although the model trained on datA does not generalize well to datB, a multiple linear regression model refitted on datB remains statistically significant (p < 0.001) and achieves a similar level of predictive performance when evaluated on its own test set (RMSE for datA = `r lmA_rmse`, for datB = `r lmB_rmse`).

This suggests that the overall prediction difficulty is comparable across the two datasets, while the predictor–outcome relationships differ between them.

## *○ Changed in prediction*  
We summarize predictors that exhibit noticeable changes in coefficient estimates between the two cohorts, including changes in direction, magnitude, or statistical significance.  
```{r}
sig_code <- function(p) {
  case_when(
    p < 0.001 ~ "***",
    p < 0.01  ~ "**",
    p < 0.05  ~ "*",
    p < 0.1   ~ ".",
    TRUE      ~ "Not sig"
  )
}

tidy_A <- tidy(lm_fit_datA) |> 
  select(term, estimate, p.value) |>
  rename(coef_A = estimate, p_A = p.value)

tidy_B <- tidy(lm_fit_datB) |>
  select(term, estimate, p.value) |>
  rename(coef_B = estimate, p_B = p.value)

compare_summary <- left_join(tidy_A, tidy_B, by = "term") |>
  filter(term != "(Intercept)") |>
  mutate(
    direction = case_when(
      sign(coef_A) != sign(coef_B) ~ "Reversed",
      TRUE ~ "Same"
    ),
    diff_num = coef_B - coef_A,
    diff = case_when(
      diff_num > 0 ~ paste0("+", round(diff_num, 4)),
      diff_num < 0 ~ as.character(round(diff_num, 4)),
      TRUE ~ "Similar"
    ),
    sig_change = paste(sig_code(p_A), "to", sig_code(p_B))
  ) |>
  select(term, coef_A, coef_B, diff, direction, sig_change)

kbl_compare <- kbl(compare_summary, digits = 4,
                   caption = "Coefficient Comparison: datA vs datB") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_compare

```

Overall, notable differences are observed between the two cohorts. Several predictors exhibit changes in the direction and magnitude of their effects, and some variables also show shifts in statistical significance. Detailed interpretation of these findings is provided in the main report.

```{r}
# Compare key variable distributions between cohorts
dist_compare <- data.frame(
  Variable = c("tumor_size (mean)", "tumor_size (sd)",
               "HER2 positive (%)", "hormonal_therapy (%)"),
  datA = round(c(mean(datA$tumor_size), sd(datA$tumor_size),
                 mean(datA$HER2 == 1) * 100,
                 mean(datA$hormonal_therapy == 1) * 100), 2),
  datB = round(c(mean(datB$tumor_size), sd(datB$tumor_size),
                 mean(datB$HER2 == 1) * 100,
                 mean(datB$hormonal_therapy == 1) * 100), 2)
)

kbl_dist <- kbl(dist_compare,
                caption = "Key Variable Distribution: datA vs datB") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
kbl_dist

```

### Magnitude of Difference
```{r}
p_mag <- compare_summary |>
  mutate(
    diff_num = coef_B - coef_A
  ) |>
  slice_max(diff_num, n = 6) |>
  ggplot(aes(x = reorder(term, diff_num), y = diff_num)) +
  geom_col() +
  coord_flip() +
  labs(x = "Predictor", y = "Coefficient change (B - A)") +
  theme_bw()
p_mag

```