---
title: "Data Science II Midterm Project"
author: "Midterm Project Group 29"
date: "2026-03-24"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: cosmo
---
```{r,warning=FALSE, message=FALSE}
library(tidyverse)
library(caret)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(corrplot)
library(glmnet)

library(mgcv)
library(earth)
```

# **Preparation and Inspection**  
## Loading Data  
```{r}
load("data/datA.RData")
load("data/datB.RData")

str(datA)
str(datB)
```

## Set variable type, dimension, and data summary
```{r}
datA <- datA |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

datB <- datB |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

dim(datA); summary(datA)
dim(datB); summary(datB)
```
## Check for Missing Data   
```{r}
colSums(is.na(datA))
colSums(is.na(datB))
```   


# **Part 1: PFS model (1990 - 1995)**  

## *○ Data Partition*

```{r}
set.seed(123)
# partition first, 0.8
datsplit <- initial_split(datA, prop = 0.8)
train <- training(datsplit) |> 
  select(-id)
test <- testing(datsplit) |> 
  select(-id)
```

### Data Visualization and featurePlot

```{r,warning=FALSE}
# visualization using train data
# continuous: age, bmi, tumor_size, lymph_nodes, Ki67
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x = train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
            y = train$log_PFS,
            plot = "scatter",
            span = 0.5,
            labels = c("Predictors", "log_PFS"),
            type = c("p", "smooth"),
            layout=c(3,2))
```

While looking at the features plots, we can observe that:

`lymph_nodes`: negative relationship with log_PFS — more positive lymph nodes, shorter PFS. The trend appears approximately linear.

`Ki67`: negative relationship with log_PFS. Higher proliferation index associated with shorter PFS, though the trend is relatively mild.

`age`: very weak relationship. Smooth curve is flat.

`bmi`: slight nonlinear, weak relationship. 

`tumor_size`: ngeative relationship. larger tumor size associated with shorter PFS.

As `bmi` and `tumor_size` shows the possibility of nonlinear relationships, a GAM was employed to evaluate whether nonlinear effects should be incorporated into the model.
```{r}
set.seed(123)
fit <- gam(
  log_PFS ~ s(age) + menopause + race + s(bmi) + s(tumor_size) + tumor_grade + 
    s(lymph_nodes) + ER + PR + HER2 + s(Ki67) + surgery + chemo + hormonal_therapy,
  data = datA,
  method = "REML")
summary(fit)
```
The GAM analysis indicated that BMI exhibited clear nonlinear effects, while tumor size showed mild nonlinearity. Therefore, flexible models such as GAM and MARS were considered as candidate models.

```{r}
# also take a look at the Pearson correlation coefficient
cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")], 
    train$log_PFS)
```
```{r}
# corrplot, none r<0.8
cor_matrix <- cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")])
corrplot(cor_matrix, method = "number", type = "upper")
```

Tumor size and lymph node count showed moderate-to-high correlation (r ≈ 0.70), suggesting potential multicollinearity. However, both variables are clinically meaningful and were retained in the model.

```{r}
# categorical: stage, tumor_grade, ER, PR, HER2, menopause, surgery, chemo, hormonal_therapy
# I'm using box plot instead of the feature plot
p1 <- ggplot(train, aes(x = stage, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p2 <- ggplot(train, aes(x = tumor_grade, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p3 <- ggplot(train, aes(x = ER, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p4 <- ggplot(train, aes(x = PR, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p5 <- ggplot(train, aes(x = HER2, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p6 <- ggplot(train, aes(x = menopause, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p7 <- ggplot(train, aes(x = surgery, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p8 <- ggplot(train, aes(x = chemo, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p9 <- ggplot(train, aes(x = hormonal_therapy, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

Notice that:

`stage`: Stage 1 patients having the highest log_PFS and Stage 3 the lowest (F = 81.0, p < 0.001), more advanced disease stage is associated with shorter PFS.

`tumor_grade`: Grade 1 patients had notably higher log_PFS, poorer tumor differentiation is associated with worse prognosis.

`ER`: ER-positive patients had substantially higher log_PFS than ER-negative patients (F = 138.4, p < 0.001)

`hormonal_therapy`: Patients receiving hormonal therapy had markedly higher log_PFS (F = 89.6, p < 0.001)

### ANOVA  
```{r}
cat_vars <- c("stage", "tumor_grade", "ER", "PR", "HER2", 
              "menopause", "surgery", "chemo", "hormonal_therapy")

anova_results <- data.frame(
  Variable = cat_vars,
  F_value = NA,
  P_value = NA
)

for(i in seq_along(cat_vars)){
  fit <- aov(log_PFS ~ train[[cat_vars[i]]], data = train)
  f_val <- summary(fit)[[1]]$`F value`[1]
  p_val <- summary(fit)[[1]]$`Pr(>F)`[1]
  anova_results$F_value[i] <- round(f_val, 3)
  anova_results$P_value[i] <- round(p_val, 4)
}

anova_results
```

## *○ Model Selection*   
Since the response variable, log-transformed PFS, is continuous, a linear regression model was considered appropriate. Exploratory data analysis showed that most continuous predictors had approximately linear relationships with log_PFS, with only mild non-linearity observed for BMI. Therefore, linear regression was used as the primary modeling approach.

In addition, given the evidence of nonlinear effects between log(PFS) and BMI, flexible models such as GAM and MARS were also considered as candidate models.

Boxplots of categorical variables also suggested differences in log_PFS across groups, indicating their potential relevance in the model.

### Base model (linear)   
```{r, linear regression 1}
set.seed(123)

lm_fit <- lm(log_PFS ~ . , data = train)
summary(lm_fit)
```
The multiple linear regression model was statistically significant (p < 0.001), with an R² of `r summary(lm_fit)$r.squared` and adjusted R² of `r summary(lm_fit)$adj.r.squared`, indicating that approximately `r round(summary(lm_fit)$r.squared,2)*100`% of the variability in log-transformed PFS was explained by the predictors.

### Resampling and Alternative models   
Considering the potential limitations of linear models in predictive performance, and given that exploratory analysis suggested the presence of nonlinear relationships among some variables, more flexible models such as GAM and MARS were introduced to capture potential nonlinear effects. 

In addition, given the possible correlations among predictors, regularization methods including Ridge, LASSO, and Elastic Net were further considered to improve model stability and generalization. 

Finally, all candidate models were evaluated and compared using cross-validated RMSE implemented via the caret package.

#### Cross-validation setup
```{r, Cross validation setup}
set.seed(123)
folds <- createMultiFolds(train$log_PFS, k = 10, times = 5)

ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  index = folds
)
```

#### Linear Model and Regularized Models
```{r, Linear}
set.seed(123)
# LM

lm_fit <- train(log_PFS ~ ., data = train, method = "lm", metric = "RMSE", trControl = ctrl)

# Ridge

ridge_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 0, lambda = exp(seq(-5, 5, length = 100))),
                   trControl = ctrl)

# Lasso

lasso_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                   tuneGrid = expand.grid(alpha = 1, lambda = exp(seq(-5, 2, length = 100))),
                   trControl = ctrl)

# Elastic Net

enet_fit <- train(log_PFS ~ ., data = train, method = "glmnet", metric = "RMSE",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 21),
                                         lambda = exp(seq(-5, 5, length = 100))),
                  trControl = ctrl)
```

#### Nonlinear Models
```{r, Nonlinear}
set.seed(123)
# GAM 

gam_fit <- train(log_PFS ~ ., data = train, method = "gam",  metric = "RMSE", trControl = ctrl)

# MARS

mars_fit <- train(log_PFS ~ ., data = train, method = "earth", metric = "RMSE",
                  tuneGrid = expand.grid(degree = 1:3, nprune = 2:20),
                  trControl = ctrl)
```

### RMSE results
```{r}
resamp <- resamples(list(LM = lm_fit, Ridge = ridge_fit, Lasso = lasso_fit,
                         ENet = enet_fit, GAM = gam_fit, MARS = mars_fit))
summary(resamp)
bwplot(resamp, metric = "RMSE")
```

The predictive performance of all candidate models was highly comparable, with only minimal differences in RMSE. Although GAM achieved the lowest RMSE, the improvement over the linear model was negligible (<1%), suggesting that increased model flexibility did not lead to meaningful gains in predictive performance, the linear regression model already captures most of the signal. 

## *○ Final model*   
Given the similar performance and the advantage of interpretability, the linear model was selected as the final model.
```{r}
set.seed(123)

lm_pred <- predict(lm_fit, newdata = test)

lm_test_rmse <- sqrt(mean((test$log_PFS - lm_pred)^2))
lm_test_r2 <- cor(test$log_PFS, lm_pred)^2

lm_test_rmse
lm_test_r2
```