---
title: "Data Science II Midterm Project"
author: "Midterm Project Group 29"
date: "2026-03-24"
output: 
  html_document:
    theme: cosmo
---
```{r,warning=FALSE}
library(tidyverse)
library(caret)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(corrplot)

set.seed(123)
```
# Loading and Cleaning data
```{r}
load("data/datA.RData")
load("data/datB.RData")

str(datA)
str(datB)
```

```{r}
datA <- datA |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

datB <- datB |> 
  mutate(
    across(c(age, bmi, tumor_size, lymph_nodes, Ki67), as.numeric),
    across(c(menopause, ER, PR, HER2, surgery, chemo, hormonal_therapy), factor),
    race = factor(race),
    tumor_grade = factor(tumor_grade),
    stage = factor(stage)
  )

dim(datA); summary(datA)
dim(datB); summary(datB)
```

```{r}
colSums(is.na(datA))
colSums(is.na(datB)) # Check if any data missing.
```
# Data Partition
```{r}
# partition first, 0.8
datsplit <- initial_split(datA, prop = 0.8)
train <- training(datsplit)
test <- testing(datsplit)
```

# Data Visualization and featurePlot
```{r,warning=FALSE}
# visualization using train data
# continuous: age, bmi, tumor_size, lymph_nodes, Ki67
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x = train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")],
            y = train$log_PFS,
            plot = "scatter",
            span = .5,
            labels = c("Predictors", "log_PFS"),
            type = c("p", "smooth"),
            layout = c(3, 2))
```

While looking at the features plots, we can observe that:

`lymph_nodes`: negative relationship with log_PFS — more positive lymph nodes, shorter PFS. The trend appears approximately linear.

`Ki67`: negative relationship with log_PFS. Higher proliferation index associated with shorter PFS, though the trend is relatively mild.

`age`: very weak relationship. Smooth curve is flat.

`bmi`: slight negative relationship. 

`tumor_size`: ngeative relationship. larger tumor size associated with shorter PFS.

```{r}
# also take a look at the Pearson correlation coefficient
cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")], 
    train$log_PFS)
```
```{r}
# corrplot, none r<0.8
cor_matrix <- cor(train[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67")])
corrplot(cor_matrix, method = "number", type = "upper")
```

```{r}
# categorical: stage, tumor_grade, ER, PR, HER2, menopause, surgery, chemo, hormonal_therapy
# I'm using box plot instead of the feature plot
p1 <- ggplot(train, aes(x = stage, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p2 <- ggplot(train, aes(x = tumor_grade, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p3 <- ggplot(train, aes(x = ER, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p4 <- ggplot(train, aes(x = PR, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p5 <- ggplot(train, aes(x = HER2, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p6 <- ggplot(train, aes(x = menopause, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p7 <- ggplot(train, aes(x = surgery, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p8 <- ggplot(train, aes(x = chemo, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
p9 <- ggplot(train, aes(x = hormonal_therapy, y = log_PFS)) + 
      geom_boxplot() + theme_bw()
(p1 + p2 + p3) / (p4 + p5 + p6) / (p7 + p8 + p9)
```

Notice that:

`stage`: Stage 1 patients having the highest log_PFS and Stage 3 the lowest (F = 81.0, p < 0.001), more advanced disease stage is associated with shorter PFS.

`tumor_grade`: Grade 1 patients had notably higher log_PFS, poorer tumor differentiation is associated with worse prognosis.

`ER`: ER-positive patients had substantially higher log_PFS than ER-negative patients (F = 138.4, p < 0.001)

`hormonal_therapy`: Patients receiving hormonal therapy had markedly higher log_PFS (F = 89.6, p < 0.001)

```{r}
cat_vars <- c("stage", "tumor_grade", "ER", "PR", "HER2", 
              "menopause", "surgery", "chemo", "hormonal_therapy")

anova_results <- data.frame(
  Variable = cat_vars,
  F_value = NA,
  P_value = NA
)

for(i in seq_along(cat_vars)){
  fit <- aov(log_PFS ~ train[[cat_vars[i]]], data = train)
  f_val <- summary(fit)[[1]]$`F value`[1]
  p_val <- summary(fit)[[1]]$`Pr(>F)`[1]
  anova_results$F_value[i] <- round(f_val, 3)
  anova_results$P_value[i] <- round(p_val, 4)
}

anova_results
```

